---
title: "Zotu_map"
author: "Michael Mann"
date: "5/9/2022"
output: html_document
---

First cleaning up tedersoo table and rarefying it. 

Then I clean up zotu table for arctic samples. 

```{r setup}
library(tidyverse)
library(ggmap)
library(future)
#library(EcolUtils)
```

first i will process Tedersoos dataset and make it easier to work with
```{r read in data}


Fungi_GSMc_OTU_Table <- 
  read_table("Tedersoo_dataset/Fungi_GSMc_OTU_Table.txt")
```



```{r functions needed}


# used to drop samples with too few reads
sample_depth_calc <- function(data){
  sum(data$Abundance)
}

#create custom function that doen't print out which permutation is completed. 
#library(EcolUtils) used for rrarefy.perm() for computing 1000 rarefactions
#these create a vector of the samples that pass
rrarefy.perm_custom <- function (x, sample = min(rowSums(x)), n = 100, round.out = T){
  require(vegan)
  y <- rrarefy(x, sample)
  for (i in 2:n) {
    #cat("Permutation ", i, " out of ", n, "\n")
    y <- y + rrarefy(x, sample)
  }
  if (round.out == T) 
    y <- round(y/n)
  if (round.out == F) 
    y <- y/n
  y
}



# just pivot and make remove zeros
cleanup <- function(data){
  data %>%
  pivot_longer(cols = where(is.numeric), names_to = "Sample", values_to = "Abundance") %>% 
  filter(Abundance > 0)
  
}

### main function ### 
# run every rarefaction 1000 times and sample at the depth provided (1500)
cleanup_rrarefy_perm  <- function(input, depth){
input %>%
  group_by(Sample) %>%
  nest() %>%
  ungroup() %>%
  mutate(sample_depth = map_dbl(data, sample_depth_calc)) %>% # calculate reads per sample
  filter(sample_depth >= depth) %>% # drop samples with fewer than 1500 reads
  select(-sample_depth) %>%
  unnest(data) %>%
  pivot_wider(names_from = OTU, values_from = Abundance, values_fill = 0) %>% # make it wide again
  column_to_rownames("Sample") %>%
  rrarefy.perm_custom(n = 1000, sample = depth) %>%
  as.data.frame() %>%
  rownames_to_column("Sample") %>%
  pivot_longer(cols = where(is.numeric), names_to = "OTU", values_to = "Abundance") %>%
  filter(Abundance > 0)
}






```

```{r splits}
# no memory for running everything at once so i split up the function into three smaller jobs. 
# splitting up the table into 8 batches.
Fungi_GSMc_OTU_Table_raw <- 
  # create all the splits by subsetting
  tibble(split = list(select(Fungi_GSMc_OTU_Table, c(1:400)), 
                      select(Fungi_GSMc_OTU_Table, c(1, 401:800)),
                      select(Fungi_GSMc_OTU_Table, c(1, 801:1200)),
                      select(Fungi_GSMc_OTU_Table, c(1, 1201:1600)),
                      select(Fungi_GSMc_OTU_Table, c(1, 1601:2000)),
                      select(Fungi_GSMc_OTU_Table, c(1, 2001:2400)),
                      select(Fungi_GSMc_OTU_Table, c(1, 2401:2800)),
                      select(Fungi_GSMc_OTU_Table, c(1, 2801:3201)))) %>%
  mutate(raw = map(split, .f = cleanup)) %>% # run the cleanup function that pivots and removes zeros
  select(-split) # drop intial data

# drop the original table
rm(Fungi_GSMc_OTU_Table)





# essentially rarefy but way faster (i drop things below 1 )
depth <- 1000

Fungi_GSMc_OTU_Table_rarefied <- 
  Fungi_GSMc_OTU_Table_raw %>%
  unnest(raw) %>%
  group_by(Sample) %>%
  nest() %>%
  ungroup() %>%
  mutate(sample_depth = map_dbl(data, sample_depth_calc)) %>% # calculate reads per sample
  filter(sample_depth >= depth) %>% # drop samples with fewer than 1500 reads
  select(-sample_depth) %>%
  unnest(data) %>%
  group_by(Sample) %>%
  mutate(rarefied = depth * Abundance / sum(Abundance)) %>%
  ungroup() %>%
  filter(rarefied >= 1) %>%
  select(-Abundance)



# i want to have everything merged together so its easier to switch out the data later. 

Fungi_GSMc_OTU_Table_finished <- 
  Fungi_GSMc_OTU_Table_raw %>%
  unnest(raw) %>%
  full_join(Fungi_GSMc_OTU_Table_rarefied, by = c("Sample", "OTU")) %>%
  replace_na(list(rarefied = 0))

```




doing Arctic dataset now
```{r read in arctic data}

zotu_arctic_raw <- 
  read_table("ZOTUs/zotu_table.txt") %>%
  rename(ZOTU =  `#OTU`) %>%
  pivot_longer(cols = where(is.numeric), names_to = "Sample", values_to = "Abundance") %>%
  select(-negcontrol) %>%
  filter(Abundance > 0)
  

# still using depth of 1500

zotu_arctic_rarefied <- 
  zotu_arctic_raw %>%
  group_by(Sample) %>%
  nest() %>%
  ungroup() %>%
  mutate(sample_depth = map_dbl(data, sample_depth_calc)) %>% # calculate reads per sample
  filter(sample_depth >= depth) %>% # drop samples with fewer than 1500 reads
  select(-sample_depth) %>%
  unnest(data) %>%
  group_by(Sample) %>%
  mutate(rarefied = depth * Abundance / sum(Abundance)) %>%
  ungroup() %>%
  filter(rarefied >= 1) %>%
  select(-Abundance)

### USING THIS ONE ####
zotu_arctic_finished <- 
  zotu_arctic_raw %>%
  full_join(zotu_arctic_rarefied, by = c("Sample", "ZOTU")) %>%
  replace_na(list(rarefied = 0))

```



reading in metadat so i can match up the two datasets. 
```{r adding metadata}

## Arctic metadata ##
arctic_metadata <- read_csv("Arctic_dataset/DNA_samples_CODES_old_new.csv")

arctic_coords <- read_csv("Arctic_dataset/GPS_coord_rough.csv")


arctic_taxonomy <- 
  read_table("ZOTUs/zotu_taxonomy.txt", col_names = c("ZOTU", "Stats", "Direction", "Taxonomy")) %>%
  separate(Taxonomy, into = c("kingdom", "phylum", "class", "order","family", "genus", "species"), sep = ",") %>% # split the tax out
  select(-Stats) %>%
  mutate(across(kingdom:species, ~str_remove(string = .x, pattern = "^[:alpha:]:"))) %>% # remove nonsense 
  mutate(ZOTU = str_remove(ZOTU, pattern = ";uniq.*"))

arctic_metadata <- 
  arctic_metadata %>%
  left_join(arctic_coords, by = "Location" ) %>%
  rename(Sample= `Suggest final name`)






### Tedersoo metadata ##

Fungi_GSMc_OTU_metadata <- read_table("Tedersoo_dataset/Tedersoo L, Mikryukov V, Anslan S et al. Fungi_GSMc_sample_metadata.txt")

Fungi_GSMc_OTU_taxonomy <- 
  read_csv("Tedersoo_dataset/Tedersoo L, Mikryukov V, Anslan S et al. Fungi_GSMc_taxonomy-function table_final.csv") %>%
  select(OTU:UNITE_SH)

```



Merging metadata with abundance data for arctic and tedersoo dataset. 
```{r merge metadata and abundance data}

zotu_arctic_metadata_merged <- 
  zotu_arctic_finished %>%
  left_join(arctic_metadata, by = c("Sample")) %>%
  group_by(ZOTU, Location, Latitude, Longitude) %>%
  summarise(total_raw_abundance = sum(Abundance), 
            total_rarefied_abundnace = sum(rarefied),  .groups = "drop") %>%
  drop_na(Location) %>% # didnt match up with metadata 
  mutate(summary_zotu = case_when(total_raw_abundance > 0 & total_rarefied_abundnace > 0 ~ "Common_Rarefied", 
                                  total_raw_abundance > 0 ~ "Only_Raw", 
                                  TRUE ~ "Missing" ))  %>%
  left_join(arctic_taxonomy, by = "ZOTU")

Fungi_GSMc_OTU_metadata_merged <- 
Fungi_GSMc_OTU_Table_finished %>%
  left_join(Fungi_GSMc_OTU_metadata, by = c("Sample" = "plot")) %>%
  group_by(OTU, Sample, latitude, longitude) %>%
  summarise(total_raw_abundance = sum(Abundance), 
            total_rarefied_abundnace = sum(rarefied), .groups = "drop") %>%
  mutate(summary_zotu = case_when(total_raw_abundance > 0 & total_rarefied_abundnace > 0 ~ "Common_Rarefied", 
                                  total_raw_abundance > 0 ~ "Only_Raw", 
                                  TRUE ~ "Missing" ))  %>%
  left_join(Fungi_GSMc_OTU_taxonomy, by = "OTU")
  

```



HERE Is where I can match up the data from both projects!!!
Going to create a summary file that lists the zotu and otu and a code depending on the hit. 
```{r matching zotu wihh otus}

tedersoo_hits <- read_table("ZOTUs/zotus_duplicate_hits_tedersoo.b6", 
                   col_names = c("Query_Label", "Target_Label", "OTU_Taxonomy",
                                 "Percent_Identity", "Alignment_Length", 
                                 "Mismatches", "Gaps", "Start_Query", "End_Query", "Start_Target", "End_Target", 
                                 "E_Value", "Bit_score"))

#### IMportant file ####
# matches up zotus and otus!
matching_hits <- 
  tedersoo_hits %>%
  rename(ZOTU = Query_Label, 
         OTU = Target_Label) %>%
  mutate(ZOTU = str_remove(ZOTU, pattern = ";uniq.*")) %>% # just so it matches the rest
  select(-OTU_Taxonomy, -E_Value, -Bit_score) %>%
  mutate(result = case_when(Percent_Identity == 100 ~ "Identical", 
                            Percent_Identity >= 99 ~ "99% match", 
                             Percent_Identity >= 97 ~ "97% match")) 
```


now I have all the pieces!

Now to merge two datasets. 


going to use a tacomic depth (family or genus)
```{r merging two datasets}
ZOTU_list <- 
  zotu_arctic_metadata_merged %>%
  group_by(ZOTU) %>%
  summarise(rarefied_total = sum(total_rarefied_abundnace), 
            abudance_total = sum(total_raw_abundance)) %>%
  arrange(desc(rarefied_total), desc(abudance_total)) %>%
  filter(rarefied_total > 0) %>%
  pull(ZOTU)
  

for(i in 1:length(ZOTU_list)){

    zotu <- ZOTU_list[i]
    
    zotu_genus <- 
      arctic_taxonomy %>%
      filter(ZOTU == zotu) %>%
      pull(genus)
    
    
    zotu_family <- 
      arctic_taxonomy %>%
      filter(ZOTU == zotu) %>%
      pull(family)
    
    
    common_arctic <- 
      zotu_arctic_metadata_merged %>%
      filter(ZOTU == zotu) %>%
      select(ZOTU:Longitude)
    
    world_matched <- 
      Fungi_GSMc_OTU_metadata_merged %>%
      left_join(matching_hits, by = "OTU")
    
    priortize_hits <- function(data){
      point_system <- tibble( result = c("97% match", "99% match", "Identical"), point = c(1, 2, 3))
      
      data %>%
        left_join(point_system, by = "result") %>%
        arrange(desc(point)) %>%
        slice(1)
    }
    
    
  
    world_hits <- 
      world_matched %>%
      rename(summary_otu = summary_zotu) %>%
      mutate(result = factor(result, levels = c("97% match", "99% match", "Identical"))) %>%
      mutate(summary_otu = factor(summary_otu, levels = c("Common_Rarefied", "Only_Raw"))) %>%
      filter(ZOTU == zotu) %>%
      group_by(Sample) %>%
      nest() %>%
      mutate(data = map(data, priortize_hits)) %>% 
      unnest(data) 
    
    world_genus <- 
      Fungi_GSMc_OTU_metadata_merged %>%
      filter(genus == zotu_genus) 
  
    
    ### IF THERE ARE NO HITS TO IT. FOLLOW THIS part
    if (nrow(world_hits) == 0){
      zotu_plot <- 
      world_map %>%
      ggplot(aes(x = long, y = lat)) +
      geom_polygon(aes(group = group), fill="lightgray") + 
     # geom_point(data = world_hits, aes(x = longitude, y = latitude, color = result, shape = summary_otu),  size = 0.3, alpha = .5) +
      geom_point(data = common_arctic, aes(x = Longitude, y = Latitude), size = 0.1, color = "purple", alpha = .5) +
      ggtitle(paste0(zotu, ", ", zotu_genus)) + scale_color_manual(values = color_scheme)+  scale_shape_manual(values = shape_scheme) + 
      guides(color = guide_legend(override.aes = list(size = 2)),
             shape = guide_legend(override.aes = list(size = 2)))
      
    
    genus_plot <- 
      world_map %>%
      ggplot(aes(x = long, y = lat)) +
      geom_polygon(aes(group = group), fill="lightgray") + 
      geom_point(data = world_genus, aes(x = longitude, y = latitude), color = "black", alpha = .1, size = 0.3)  + ggtitle(zotu_genus)
    
    
    
    
    combined_fig <- cowplot::plot_grid(zotu_plot + theme(legend.position="none"), genus_plot, ncol = 1)
    
    # create folders for output
    dir.create(file.path(getwd(), "maps",  zotu_family,  zotu_genus), showWarnings = FALSE, recursive = TRUE)
    
    
    
    
    file_path <- paste0(file.path(getwd(), "maps", zotu_family,  zotu_genus), "/", zotu, "_plot.pdf")
    
    
    cowplot::save_plot(file_path, combined_fig, base_asp = 1.1)
      
      
    } else {
    
    
      
    
    color_scheme <- setNames( c('red', 'forestgreen', 'blue'), levels(world_hits$result)  )
    shape_scheme <- setNames( c(3, 2), levels(world_hits$summary_otu)  )
    
    world_genus <- 
      Fungi_GSMc_OTU_metadata_merged %>%
      filter(genus == zotu_genus) 
    
    world_map <- map_data("world")
    
    
    zotu_plot <- 
      world_map %>%
      ggplot(aes(x = long, y = lat)) +
      geom_polygon(aes(group = group), fill="lightgray") + 
      geom_point(data = world_hits, aes(x = longitude, y = latitude, color = result, shape = summary_otu),  size = 0.3, alpha = .5) +
      geom_point(data = common_arctic, aes(x = Longitude, y = Latitude), size = 0.1, color = "purple", alpha = .5) +
      ggtitle(paste0(zotu, ", ", zotu_genus)) + scale_color_manual(values = color_scheme)+  scale_shape_manual(values = shape_scheme) + 
      guides(color = guide_legend(override.aes = list(size = 2)),
             shape = guide_legend(override.aes = list(size = 2)))
      
    
    genus_plot <- 
      world_map %>%
      ggplot(aes(x = long, y = lat)) +
      geom_polygon(aes(group = group), fill="lightgray") + 
      geom_point(data = world_genus, aes(x = longitude, y = latitude), color = "black", alpha = .1, size = 0.3)  + ggtitle(zotu_genus)
    
    
    legend <- 
        cowplot::get_legend(
        zotu_plot + 
        guides(color = guide_legend(nrow = 1)) +
        theme(legend.position = "bottom"))
    
    
    
    combined_fig <- cowplot::plot_grid(zotu_plot + theme(legend.position="none"), genus_plot, ncol = 1)
    
    # create folders for output
    dir.create(file.path(getwd(), "maps",  zotu_family,  zotu_genus), showWarnings = FALSE, recursive = TRUE)
    
    
    
    
    file_path <- paste0(file.path(getwd(), "maps", zotu_family,  zotu_genus), "/", zotu, "_plot.pdf")
    
    
    cowplot::save_plot(file_path, combined_fig, base_asp = 1.1)
    }
  }


  cowplot::save_plot(paste0(file.path(getwd(), "maps/legend.pdf")), legend)

```




